A novel framework for linking dynamic experiences to memories reveals event-like structure in naturalistic episodic recall

### 2019-07-21 paper/main.tex
saving progress

\begin{abstract}
Our life experiences unfold over time and the temporal dynamics of their contents form unique \textit{experience trajectories}.  Within this geometric framework, one can compare the shape of the trajectory formed by an experience to that defined by our later remembering of that experience.  We propose a framework for mapping naturalistic experiences onto geometric spaces that characterize how they unfold over time.  We apply this approach to a naturalistic memory experiment which had participants view and recount a video.  We find that the video and subsequent recall share an event-like structure, and that the ``shapes'' of the trajectories formed by participants' recountings were all highly similar to that of the original video. Further, participants' that recounted the events with a high level of precision and in a distinctive way had better subsequent memory performance. Lastly, we identified a network of brain structures that are sensitive to the shapes of our ongoing experiences, and an overlapping network that is sensitive to how we will later remember those experiences. These results highlight the rich event-like structure of memories for our life experiences, and introduce a novel framework for mapping between dynamic life experiences and their mnemonic counterparts.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Using well-controlled trial-based experimental designs, the field has amassed a wealth of valuable information regarding human episodic memory.  However, there are fundamental properties of our memories that trial-based experiments are not well suited to capture~\citep[for review also see][]{KoriGold94, HukEtal18}.  First, our memories are continuous, rather than discrete: removing a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific language used to describe an experience have little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, counting the number or proportion of precise recalls is often a primary metric of assessing the quality of participants' memories.  Third, one might remember the \textit{essence} (or shape) of an experience but forget (or neglect to recount) particular details.  Capturing the essence of what happened is typically the main ``point'' of recounting a memory to a listener.

How might one go about formally characterizing the shape of an experience, or whether that shape has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g., ][]{LernEtal11}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can then leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

Although our experiences most often change gradually, they also occasionally change suddenly~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) during an experience help to discretize our experiences into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within event) and transient (across event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distill those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}. Prior research has implicated the hippocampus and the medial prefrontal cortex as playing a critical role in transforming experiences into stuctured and consolidated memories ~\citep{TompDava17}.

Here we sought to examine how the temporal dynamics of a ``naturalistic'' experience were reflected in participants' later memories of that experience.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recalled an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and we use Hidden Markov Models~\citep{Rabi89, BaldEtal17} to discretize the evolving semantic content into events.  In this way, we cast naturalistic experiences (and recalls of those experiences) as \textit{trajectories} that describe how the experiences evolve over time. Under this framework, successful remembering entails verbally ``traversing'' the topic trajectory of the original episode, thereby reproducing the shape of the original experience.  Comparing the shapes of the topic trajectories of the original episode and of participants' retellings of the episode reveals which aspects of the episode were preserved (or lost) in the translation into memory.

Here, we introduce a novel content-based framework for linking dynamic experiences to memories. We hypothesized that the ``shape'' and the event-like structure of an experience would be preserved in memory. Further, we tested  1) whether the \textit{precision} with which the participant recounts each event and 2) how \textit{distinctive} each recall event is (relative to the other recalled events) relates to overall memory performance. Last, we identify networks of  brain structures whose responses (as participants watched the episode) reflected the shape of the episode, and how participants would later recount the episode.

\section*{Results}
To characterize the shape of the \textit{Sherlock} episode participants watched and their subsequent recountings of the episode, we used a topic model~\citep{BleiEtal03} to discover the latent thematic content in the video.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents; they return as output two matrices.  The first output is a \textit{topics matrix} whose rows are topics (latent themes) and whose columns correspond to words in the vocabulary. The entries of the topics matrix define how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix}, with one row per document and one column per topic.  The topic proportions matrix describes which mixture of topics is reflected in each document.

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) scenes spanning the roughly 45 minute video used in their experiment.  This information included: a brief narrative description of what was happening; whether the scene took place indoors vs. outdoors; names of any characters on the screen; names of any characters who were in focus in the camera shot; names of characters who were speaking; the location where the scene took place; the camera angle (close up, medium, long, etc.); whether or not background music was present; and other similar details (for a full list of annotated features see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping 50-scene sliding windows, and treated each 50-scene sequence as a single ``document'' for the purposes of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 27 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the video (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06}, in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each timepoint of the episode (we set timepoints to match the acquisition times of the 1976 fMRI volumes collected as participants viewed the episode).

@@

The topics we found were heavily character-focused (e.g., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes that were primarily Sherlock Holmes-focused (Sherlock is the titular character); primarily John Watson-focused (John is Sherlock's close confidant and assistant); or that involved Sherlock and John interacting (Fig.~\topics).  Several of the topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences (if the distinctions between those overlapping topics were meaningful; also see Fig.~\featureimportance).  The topic vectors for each timepoint were \textit{sparse}, in that only a small number (usually one or two) of topics tended to be ``active'' in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistance} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B).  Following \cite{BaldEtal17}, we used a Hidden Markov Model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., at the boundaries of the blocks in the correlation matrix; event boundaries identified by the HMM are outlined in yellow).  Part of our model fitting procedure required selecting an appropriate number of ``events'' to segment the timeseries into.  We used an optimization procedure to identify the number of events that maximized within-event stability while also minimizing across-event correlations (see \textit{Methods} for additional details).  To create a stable ``summary'' of the video, we computed the average topic vector within each event (Fig.~\ref{fig:model}C).

>>>>

\begin{abstract}
Our life experiences unfold over time in highly complex manner, with the evolving presence and absence of numerous intricate features describing our journey between each circumstance or event we encounter. Here, we propose a framework for mapping dynamic naturalistic experiences onto geometric spaces as \textit{trajectories} that capture the temporal dynamics of real-world content. Within this geometric framework, one may compare the shape of the trajectory formed by an experience to that defined by one's later recollection to characterize our memories' recovery and distortion of the external world. Here, we apply this approach to a naturalistic memory experiment in which participants viewed and verbally recounted a video, and find that the video and subsequent recalls share both an experience-specific shape and a discernible event-like structure.  However, the level of \textit{precision} with which individuals recounted various events and the \textit{distinctiveness} of recall for those events were varied and predictive of overall memory performance. Finally, we identify a network of brain structures that is sensitive to the ``shapes" of our ongoing experiences, and an overlapping network sensitive to how we will later remember them.  These results highlight the rich event-like structure of the external world and our memories, and offer novel, content-sensitive alternatives to classical ``proportion recalled" measures for assessing episodic memory
%Our life experiences unfold over time and the temporal dynamics of their contents form unique \textit{experience trajectories}.  Within this geometric framework, one can compare the shape of the trajectory formed by an experience to that defined by our later remembering of that experience.  We propose a framework for mapping naturalistic experiences onto geometric spaces that characterize how they unfold over time.  We apply this approach to a naturalistic memory experiment which had participants view and recount a video.  We find that the video and subsequent recall share an event-like structure, and that the ``shapes'' of the trajectories formed by participants' recountings were all highly similar to that of the original video. Further, participants' that recounted the events with a high level of precision and in a distinctive way had better subsequent memory performance. Lastly, we identified a network of brain structures that are sensitive to the shapes of our ongoing experiences, and an overlapping network that is sensitive to how we will later remember those experiences. These results highlight the rich event-like structure of memories for our life experiences, and introduce a novel framework for mapping between dynamic life experiences and their mnemonic counterparts.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of valuable information regarding human episodic memory.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep[for review also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---removing a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific language used to describe an experience has little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of precise recalls is often a primary metric for assessing the quality of participants' memories.  Third, one might remember the \textit{essence} (or a general summary) of an experience but forget (or neglect to recount) particular details.  Capturing the essence of what happened is typically the main ``point'' of recounting a memory to a listener, while the addition of highly specific details may add comparatively little to successful conveyance of an experience.

How might one go about formally characterizing the essence of an experience, or whether it has been recovered by the rememberer?  Any given moment of an experience derives meaning from surrounding moments, as well as from longer-range temporal associations~\citep[e.g., ][]{LernEtal11, Mann19}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g., ][]{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15}.  Our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we are in at any given moment is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

Although our experiences most often change gradually, they also occasionally change suddenly~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) during an experience help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within event) and transient (across event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distill those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}. Prior research has implicated the hippocampus and the medial prefrontal cortex as playing a critical role in transforming experiences into stuctured and consolidated memories ~\citep{TompDava17}.

Here we sought to examine how the temporal dynamics of a ``naturalistic'' experience were later reflected in participants' memories.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and Hidden Markov Models~\citep{Rabi89, BaldEtal17} to discretize the evolving semantic content into events.  In this way, we cast naturalistic experiences (and recalls of those experiences) as \textit{trajectories} that describe how the experiences evolve over time. Under this framework, successful remembering entails verbally ``traversing'' the content trajectory of the episode, thereby reproducing the shape (or essence) of the original experience.  Comparing the shapes of the topic trajectories of the episode and of participants' retellings of the episode then reveals which aspects of the episode were preserved (or lost) in the translation into memory.  We further examine whether 1) the \textit{precision} with which a participant recounts each event and 2) the \textit{distinctiveness} each recall event is (relative to the other recalled events) relates to their overall memory performance.  Last, we identify networks of  brain structures whose responses (as participants watched the episode) reflected the shape of the episode, and how participants would later recount the episode.

\section*{Results}
To characterize the shape of the \textit{Sherlock} episode and participants' subsequent recountings of its unfolding, we used a topic model~\citep{BleiEtal03} to discover the latent themes in the episode's dynamic content.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents; they return as output two matrices.  The first output is a \textit{topics matrix} whose rows are topics (latent themes) and whose columns correspond to words in the vocabulary. The entries of the topics matrix define how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix}, with one row per document and one column per topic.  The topic proportions matrix describes which mixture of topics is reflected in each document.

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) scenes spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening; whether the scene took place indoors vs. outdoors; names of any characters on the screen; names of any characters who were in focus in the camera shot; names of characters who were speaking; the location where the scene took place; the camera angle (close up, medium, long, etc.); whether or not background music was present; and other similar details (for a full list of annotated features see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping, 50-scene sliding windows, and treated each 50-scene sequence as a single ``document'' for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 27 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the video (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06}, in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each timepoint of the episode (we set timepoints to match the acquisition times of the 1976 fMRI volumes collected as participants viewed the episode).

@@

The topics we found were heavily character-focused (e.g., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes that were primarily Sherlock Holmes-focused (Sherlock is the titular character); primarily John Watson-focused (John is Sherlock's close confidant and assistant); or that involved Sherlock and John interacting (Fig.~\topics).  Several of the topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences (if the distinctions between those overlapping topics were meaningful; also see Fig.~\featureimportance).  The topic vectors for each timepoint were \textit{sparse}, in that only a small number (usually one or two) of topics tended to be ``active'' in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistance} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topics would appear to spring into or out of existence).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the contextual dynamics of real-world experiences.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a Hidden Markov Model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., at the boundaries of the blocks in the correlation matrix; event boundaries identified by the HMM are outlined in yellow).  Part of our model fitting procedure required selecting an appropriate number of ``events'' to segment the timeseries into.  We used an optimization procedure to identify the number of events that maximized within-event stability while also minimizing across-event correlations (see \textit{Methods} for additional details).  To create a stable ``summary'' of the video, we computed the average topic vector within each event (Fig.~\ref{fig:model}C).
